---
title: "Help_Frontiers_Ecosystem_APT"
author: "Colas Guillon"
date: "2025-01-28"
output: html_document
---

# This loads the packages
```{r, Library import}
library(terra)#package to read rasters
library(sf) #package to read vectors in a data frame format
library(ggplot2)
library(dplyr)
library(xts)
```

# Import the data 
NDVI = 230 layers
CHIRPS = 720
```{r, Open study area}
# Vectors
study_area <- vect("./Shapefile/Zone_etude_Kariba_lake.shp")
mcc <- vect("./Shapefile/MCC.shp")
protected_areas <- vect("./Shapefile/protected_areas.shp")

# Rasters
# Land use type raster for mcc region - integers for types 
# Values are NaN, 1, 3, 51, 53, 52, 7, 6, 10, 43, 42, 41, 2
lulc_mcc <-rast("./rasters/LULC_MCC.tif")

#Land use type raster for the Kariba region
# Values are 4, 2, 5, 7, 3, 1, 10, 6, NaN
lulc_kariba <- rast("./Rasters/detailled_LULC_Kariba_area.tif")

# NDVI
ndvi <- rast("./rasters/NDVI.tif")

#CHIRPS
chirps <- rast("./rasters/CHIRPS_pentad.tif")
```

# Plot the Land Use Types of MCC for last image in the stack
```{r, open rasters}
plot(chirps[[16]])

nlyr(chirps)
```

You can look at the different ndvi values in the parks, without and in the crops (do mean) and without looking the whole annual variation

# Reproject the files to match the NDVI data
```{r, miombo versus mopane on the whole kariba area}
# Reproject and take resolution of the NDVI file
lulc_kariba <- project(lulc_kariba,ndvi,method="near") #we reproject the classification but then it takes the ndvi resolution

#Reproject the study area too
study_area <- project(study_area,lulc_kariba)#We have to put in the same georeferenced system
```

#Crop and mask the rasters
This crops it to only the study area and then masks the area.
```{r}
# Crop and mask the land use file
lulc_kariba_masked <- lulc_kariba %>%
  crop(study_area) %>%
  mask(study_area)

# Same for NDVI
ndvi_masked <- ndvi %>%
  crop(study_area) %>%
  mask(study_area)

# Same for CHIRPS
chirps_masked <- chirps
```

# Isolate each of the landcover classes from the land use file
```{r}
# Extract the indices of the cells which are of each landcover type
miombo <- which(lulc_kariba_masked[]==4) #we select the type of forests
mopane <- which(lulc_kariba_masked[]==5)
riparian <- which(lulc_kariba_masked[]==7)
crop <- which(lulc_kariba_masked[]==2)
```

# Extract the NDVI for each landcover type
This extracts the NDVI value of each raster at the indices which correspond to the land cover types.
It stores each raster layer as a column and the rows are cells (in a dataframe)
```{r}
Miombo.NDVI=ndvi_masked[miombo] #we extract the ndvi values for the type of forests
Mopane.NDVI=ndvi_masked[mopane]
Riparian.NDVI=ndvi_masked[riparian]
crop.NDVI=ndvi_masked[crop]
```

#Take the mean of the ndvi for each landcover type
```{r, mean ndvi}
Miombo.NDVI.mean=colMeans(Miombo.NDVI,na.rm=T)
Mopane.NDVI.mean=colMeans(Mopane.NDVI,na.rm=T)
Riparian.NDVI.mean=colMeans(Riparian.NDVI,na.rm=T)
crop.NDVI.mean=colMeans(crop.NDVI,na.rm=T)
```

#Extract the dates from the NDVI
```{r, dates}
# This extracts the dates as a character string
ndvi.dates=names(ndvi_masked) #Extract the name of ndvi bands

# Because there the format is 2015_03_18_NDVI, we want to get rid of the NDVI
# Therefore we remove it using sub()
dates = sub("_NDVI", "", ndvi.dates) #we keep only the date

# Then we convert the character string into a date format object
dates_formattees <- as.Date(dates,"%Y_%m_%d") #put in date format
```

# Plot the Data
```{r,plot}
# Put the data into a dataframe
ndvi_mean=data.frame("dates"=dates_formattees,"Miombo"=Miombo.NDVI.mean,"Mopane"=Mopane.NDVI.mean,"Riparian"=Riparian.NDVI.mean,"Crop"=crop.NDVI.mean)#we put it in the same data frame

ndvi_long <- ndvi_mean %>% # we put it in a long format
  tidyr::pivot_longer(cols = c(Miombo, Mopane, Riparian,Crop), 
               names_to = "Type", 
               values_to = "NDVI")
ggplot(ndvi_long,aes(x=dates,y=NDVI,color=Type))+ 
  geom_line()+
  scale_color_manual(values = c("Miombo" = "green", "Mopane" = "forestgreen", "Riparian" = "lightblue","Crop"="orange")) +  
  labs(title = "Évolution du NDVI", x = "Date", y = "NDVI", color = "Végétation") + 
  theme_minimal()

```


```{r, rain}
# Changing to same resolution
chirps <- resample(chirps,lulc_kariba_masked,method="bilinear") #we need to put at the same resolution than ndvi & lulc
#Crop
chirps <- crop(chirps,study_area)
#Mask - everything outside no data
chirps_masked <- mask(chirps,study_area)

rain.miombo=chirps_masked[miombo] #we extract the rain values on miombo forests
rain.mopane=chirps_masked[mopane]
rain.riparian=chirps_masked[riparian]
rain.crop=chirps_masked[crop]
```
```{r}
# you could calculate the mean rainfall per area, per types of forest
```

```{r}
rain.dates=names(chirps_masked) #Extract the name of ndvi bands
dates = sub("_precipitation", "", rain.dates) #we keep only the date
dates_formattees <- as.Date(dates,"%Y%m%d")

# Conversion to Monthly Sum
rain.miombo.mean=colMeans(rain.miombo,na.rm=T) # Mean of rain over miombo
rain.miombo.monthly=aggregate(rain.miombo.mean,by=list(format(dates_formattees,"%Y %m")),sum,na.rm=T) %>% 
  rename("Date"=Group.1,"Rain_miombo"=x)

rain.mopane.mean=colMeans(rain.mopane,na.rm=T) 
rain.mopane.monthly=aggregate(rain.mopane.mean,by=list(format(dates_formattees,"%Y %m")),sum,na.rm=T) %>% 
  rename("Date"=Group.1,"Rain_mopane"=x)

rain.crop.mean=colMeans(rain.crop,na.rm=T) 
rain.crop.monthly=aggregate(rain.crop.mean,by=list(format(dates_formattees,"%Y %m")),sum,na.rm=T) %>% 
  rename("Date"=Group.1,"Rain_crop"=x)

rain=merge(rain.miombo.monthly,rain.mopane.monthly,by="Date")
rain=merge(rain,rain.crop.monthly,by="Date")
rain_long=tidyr::pivot_longer(rain,cols = c(Rain_miombo,Rain_mopane,Rain_crop), 
               names_to = "Type of forest", values_to = "Rain")
# Plot
ggplot(tail(rain_long,36),aes(x=Date,y=Rain,fill=`Type of forest`))+ # 36 last rows gives you 2024
  geom_col(position="dodge") #on prend que les données de 2024
```
```{r} 
#exploration de la corrélation entre pluie et ndvi
chirps_16_days <- tapp(chirps_masked, index = cut(1:nlyr(chirps_masked),length(dates_formattees)), fun = mean) # we need the rain (5 days) to fit with the ndvi (16 days)
cor(values(chirps_16_days),values(ndvi_masked),use="na.or.complete") #to dig if we want a correlation map
```


```{r}
#We can work on the different types of forest in the MCC
miombo=which(lulc_mcc[]==41 | lulc_mcc[]==42 | lulc_mcc[]==43)
mopane=which(lulc_mcc[]==51 | lulc_mcc[]==52 | lulc_mcc[]==53)
riparian=which(lulc_mcc[]==7)
```

```{r}
plot(chirps_16_days)
```

```{r}

```


